Negative Perceptions of Outsourcing to Artificial Intelligence

### 2025-06-23 manuscript.qmd
Recreate manuscript as docx and submit

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
      - editing
  - name: Jim A.C. Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: This pre-print is currently not yet peer-reviewed, and may differ from the final version.
    data-sharing: Word count (excl. Methods and Results) = 4619 words.
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but 
  also socio-relational ones. Large language models like ChatGPT can generate 
  wedding vows, speeches, and personal messages, raising questions about how 
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (N = 3,649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks. We show
  that outsourcing makes us think more negatively about not only the person and
  their motivations, but also the outsourced work itself. Moreover, we provide
  insight into why this occurs: the reduced effort from outsourcing
  socio-relational tasks to AI signals that the output is less authentically 
  one's own and that the person cares less about the task. Our research 
  highlights the way that AI use shapes our perceptions of people, raising key
  philosophical questions about efficiency, authenticity, and social ties in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, person perception, outsourcing, effort, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf: default
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

```{r}
library(kableExtra)
library(targets)
library(tidyverse)
```

The widespread release of generative AI language models has transformed daily 
life, offering the potential to perform a variety of tasks more efficiently and,
in some cases, with greater effectiveness than by doing them oneself. But as AI
becomes more widely available, people are not only using it to assist them with
things like preparing dinner recipes, writing data analysis code, and planning
daily schedules. Increasingly, AI might be used beyond routine or technical
domains to instead assist in tasks that are more socio-relational in nature,
like writing wedding vows, apology notes, and love letters. Anecdotal evidence
suggests that not only is AI-outsourcing of this kind already happening, but
that it potentially has serious effects on how we judge others. In a recent
Reddit post, a disgruntled newlywed tells the story of her husband using ChatGPT
to write his wedding vows, expressing her discomfort with outsourcing something
to AI that, to her, is deeply meaningful and a reflection of their love for one 
another [@miramar0]. Outsourcing tasks -- especially socio-relational ones -- to
AI tools may be efficient, but could have negative consequences for person 
perception.

There is nothing new, in principle, about outsourcing tasks. For hundreds of 
years, personal assistants have organised daily schedules, recipe-books have 
provided meal plans, and guidebooks have created travel itineraries. In the 
socio-relational domain, ghostwriters have long-existed, and the internet is 
abound with professional paid services for writing wedding vows and personal 
speeches. AI merely supercharges what is an ancient human impulse: the push to 
reduce mental energy by outsourcing parts of our work onto people, books, tools,
or systems. But even if outsourcing is an old phenomenon, the rapid shift in 
availability and use of AI models has fundamentally changed the ease with which
people can outsource work, what kinds of tasks they can outsource, and the way
in which they can outsource. These new developments in society mean that even as
an old phenomenon in new clothes, there is much we still need to know about
outsourcing.

First, we need to know how people who outsource tasks to AI are perceived. We 
know that people are increasingly using large language models (LLMs) for a wide
variety of tasks [@UKGov2024]. Due to their ubiquity, perhaps outsourcing to
LLMs might not lead to negative perceptions? We are sceptical. We know that
people dislike it when others "free ride" or reduce effort while benefiting from
collective resources [e.g. @Cubitt2011; @Kerr1983] and that people's outputs are
perceived as more valuable the more effort was ostensibly put into them
[@Kruger2004]. Moreover, exertion of effort is deemed morally admirable and is 
rewarded, even in situations where effort does not directly generate additional
product, quality, or economic value, suggesting that effort itself is moralised
[@Celniker2023]. Given this, even if AI tools are widely available and pitched
as improving efficiency, the core social psychological processes are likely to 
remain: someone is expending less effort to achieve a task, and people value 
effort. Indeed, some work shows that describing someone as using AI for a 
relational task led to the perception they expended less effort and were less 
satisfied with their relationship [@Liu2024] and other unpublished work looking
at perceptions of people using AI to complete academic assignments finds that 
using AI leads to more negative perceptions of moral character and suitability 
as a partner [@Roth2025].

Second, we need to know whether the *type* of task that people are outsourcing 
matters. One might expect outsourcing to be perceived negatively regardless of 
the type of task being outsourced -- if effort is generally moralised, then the 
domain in which it is expended (or not) should have little impact. However, 
there are reasons to expect differences between social tasks like writing vows 
and non-social tasks like writing computer code. We know that different norms,
standards, and expectations can be applied to social and non-social tasks and 
exchanges [e.g. @Fiske1992; @Heider1958; @Malle2022]. Moreover, from a 
philosophical perspective, it often matters not only *whether* something is 
done, but *how* it is done [@Aristotle2009; @Hursthouse2023; @Stohr2006]. An 
apology is not just about hearing someone say "I am sorry", but seeing genuine 
regret; a love letter is not just about hearing someone say "I love you", but 
seeing depth of emotion; and a bereavement letter is not just about hearing 
someone say "I am sorry for your loss", but seeing an understanding for the 
powerful human experience of loss. There is, perhaps especially for social 
tasks, value not only in the outcome of doing something, but the *process* too 
[@Goodman2010]. To understand any potential negative effects of outsourcing to 
AI, we must therefore look at a broad range of non-social and social tasks, 
rather than draw broad conclusions based on a few use cases.

Third, we need to know how different ways of outsourcing to AI influence 
negative perceptions. Someone who "fully" outsources a task to AI by simply 
giving it a prompt and copying the output word-for-word might be perceived very
differently to someone who gives the AI a prompt, revises the work accordingly,
and finishes it themselves -- using AI as a *collaborative tool*, rather than as
a replacement. Similarly, someone could deceive others about their use of AI or
be perfectly honest about it. While it seems reasonable to assume that "fully"
outsourcing would be perceived worse than using AI as a tool, and that not
acknowledging AI use would be perceived worse than being honest about it, it
remains unclear how much this reduces negative perceptions: if someone uses AI
in the "best" way, by using it as a collaborative tool and being open about this
use, would they still suffer negative social consequences from doing so?

Fourth, we need to know how outsourcing to AI, in different kinds of tasks and 
in different ways, may shape different *kinds* of social perceptions. People can
judge others on separate dimensions of warmth and competence [e.g. @Abele2021;
@Fiske2007] as well as on dimensions of morality and trustworthiness
[@Goodwin2014]. It remains unclear how outsourcing to AI might lead to 
differential character judgments across these different dimensions.

Fifth, we need to understand *why* outsourcing to AI, and therefore expending 
less effort, might have these effects. Previous work has focused on how 
expending less effort leads to negative perceptions of others [@Celniker2023].
But this raises the question of *why* effort is seen as important and what
exactly it is signalling to others, beyond one's general cooperative intent. It
is possible that outsourcing leads to negative perceptions because the lack of
effort spent on the task signals something more fundamental about how authentic
one is and how much one cares about the task: when someone chooses to outsource
a love letter to an AI, they might be seen as valuing that love letter and what
it represents less. It could be this second-step order of perceptions that is
the key driver of negative perceptions, especially for socio-relational tasks.

@@

In this paper, we build on classic social psychological work on character 
inferences from reduced effort to understand how people view others who
outsource different kinds of tasks, in different ways, for different reasons, to
AI. Across five pre-registered experiments with British participants, we seek
not only to understand how reduced effort through AI-outsourcing might shape
perceptions of others, but also to understand in more depth *why* it is that
reduced effort has the effect that it does.

In our initial pilot studies to motivate this work, we found that people who
outsource a range of tasks to AI or another person are perceived more negatively
than people who complete the tasks by themselves (see Supplementary Materials).
In Study 1, we look at the effects of task type, AI use, and honesty. We explore
how people perceive others who outsource different kinds of tasks with different
levels of social relevance (e.g., from daily schedules, computer code and dinner
recipes to wedding vows, apology letters, or bereavement cards), manipulating
whether people use AI as a collaborative tool or "fully" outsource to AI and 
whether they are honest or deceptive about their use of AI. After turning to
look at perceptions of both outsourcers and the outsourced work in Study 2, in
Studies 3-5 we probe why outsourcing may have negative effects on how we
evaluate others. In Study 3, we test potential mechanisms of perceived effort
and authenticity by looking at how people evaluate others who either spend a lot
or little time crafting the AI prompts, and who either outsource to a generic or
personalised AI. In Study 4, we test the potential mechanism of perceived
importance in the task by manipulating people's reasons for using AI -- either 
because they wanted to save time or because they cared about the task and
thought that AI would improve their work. Finally, in Study 5, we bring these 
different potential mechanisms together to explore the different pathways that 
influence the relationship between outsourcing and negative perceptions, 
focusing on perceived effort, authenticity, and care in the task.

@@

@@

We recruited a convenience sample of 800 participants from the United Kingdom 
through the online platform Prolific (<https://www.prolific.com/>). After 
excluding participants who failed our pre-treatment attention check, we were 
left with a final sample of `r nrow(study2_sample)` participants 
(`r sum(study2_sample$gender == "Female")` female; 
`r sum(study2_sample$gender == "Male")` male;
`r sum(study2_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study2_sample$gender, "Prefer") | is.na(study2_sample$gender))`
undisclosed gender; mean age = `r round(mean(study2_sample$age), 2)` years).
`r round(mean(study2_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before.

@@

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package [@Burkner2017]. We modelled each character 
evaluation -- competence, warmth, morality, laziness, and trustworthiness -- as 
a separate response variable and included fixed effects for conditions, varying 
intercepts for participants, and varying intercepts and slopes for tasks. We 
used regularising priors for all parameters to impose conservatism on parameter 
estimates. All models converged normally ($\hat{R}$ ≤ 1.01).

For all studies in this paper, we report how we determined our sample size, all
data exclusions, all manipulations, and all measures in the studies. All studies
were pre-registered. Analyses for all studies were conducted in R v4.4.2 
[@RCoreTeam]. Visualisations were produced using the *ggplot2* and *patchwork* 
packages [@Wickham2016; @Pedersen2025]. The manuscript was reproducibly 
generated using the *targets* package [@Landau2021] and *quarto* [@Allaire2024].
All data and code to reproduce the analyses and figures in this paper can be 
found here: <https://osf.io/xhmzk/?view_only=a4da193574d7410ba4d2aa3945a28b05>

@@

We first looked at the overall results averaging over tasks. Across all five 
character evaluations, we found that fully outsourcing to AI (i.e., copying the 
AI output verbatim) was perceived more negatively than using AI as a 
collaborative tool (@fig-treatments-study1; @tbl-treatment-diffs-study1). By 
contrast, we found that deception about AI usage had specific negative effects 
on perceptions of morality and trustworthiness: people who did not acknowledge 
their use of AI were perceived as less moral and less trustworthy. We did not 
find any interaction effects between full outsourcing and deception.

```{r, fig.height=6, fig.width=6}
#| label: fig-treatments-study1
#| fig-cap: Overall Character Evaluations in Study 1
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Coloured points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

>>>>

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
      - editing
  - name: Jim A.C. Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: This pre-print is currently not yet peer-reviewed, and may differ from the final version.
    data-sharing: Word count = 10,574 words.
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: As artificial intelligence (AI) tools become increasingly integrated into daily life, people are beginning to outsource not only professional tasks but also socio-relational ones. Large language models like ChatGPT can generate wedding vows, speeches, and personal messages, raising questions about how individuals who use AI for such tasks are perceived by others. In this paper, we conduct six pre-registered studies with British participants (N = 3,935) to understand how people view those who expend less effort by outsourcing tasks to AI in different ways. We show that outsourcing makes us think more negatively about not only the person and their motivations, but also the outsourced work itself. Advancing previous work on effort, we show that the specific kind of task that people outsource -- and therefore expend less effort on -- matters; that reduced effort does not lead to uniformly negative character perceptions across different tasks; and that effort is important because it signals second-order judgments that people are being less authentic and care less about the task. Our research highlights the way that AI use shapes our perceptions of people, raising key philosophical questions about efficiency, authenticity, and social ties in a world filled with AI-mediated interactions.
keywords: [artificial intelligence, person perception, outsourcing, effort, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-docx:
    fig-dpi: 300
    fig-align: center
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

```{r}
library(flextable)
library(targets)
library(tidyverse)
```

The widespread release of generative AI language models has transformed daily 
life, offering the potential to perform a variety of tasks more efficiently and,
in some cases, with greater effectiveness than by doing them oneself. But as AI
becomes more widely available, people are not only using it to assist them with
things like preparing dinner recipes, writing data analysis code, and planning
daily schedules. Increasingly, AI might be used beyond routine or technical
domains to instead assist in tasks that are more socio-relational in nature,
like writing wedding vows, apology notes, and love letters. Anecdotal evidence
suggests that not only is AI-outsourcing of this kind already happening, but
that it potentially has serious effects on how we judge others. In a recent
Reddit post, a disgruntled newlywed tells the story of her husband using ChatGPT
to write his wedding vows, expressing her discomfort with outsourcing something
to AI that, to her, is deeply meaningful and a reflection of their love for one 
another [@miramar0]. Other reports describe people's negative reactions when
they learn that their romantic partner used ChatGPT to write them an apology 
note [@Tait2024] or even to break up with them [@Anderson2025]. Outsourcing 
tasks -- especially socio-relational ones -- to AI tools may be efficient, but 
could have negative consequences for person perception.

There is nothing new, in principle, about outsourcing tasks. For hundreds of 
years, personal assistants have organized daily schedules, recipe-books have 
provided meal plans, and guidebooks have created travel itineraries. In the 
socio-relational domain, ghostwriters have long-existed, and the internet is 
abound with professional paid services for writing wedding vows and personal 
speeches. AI merely supercharges what is an ancient human impulse: the push to 
reduce mental energy by outsourcing parts of our work onto people, books, tools,
or systems. But even if outsourcing is an old phenomenon, the rapid shift in 
availability and use of AI models has fundamentally changed the ease with which
people can outsource work, what kinds of tasks they can outsource, and the way
in which they can outsource. These new developments in society mean that even as
an old phenomenon in new clothes, there is much we still need to know about
outsourcing.

First, we need to know how people who outsource tasks to AI specifically are 
perceived. We know that people are increasingly using large language models 
(LLMs) for a wide variety of tasks [@UKGov2024]. Due to their ubiquity, perhaps 
outsourcing to LLMs might not lead to negative perceptions? We are sceptical. We 
know that people dislike it when others "free ride" or reduce effort while 
benefiting from collective resources [e.g. @Cubitt2011; @Kerr1983] and that 
people's outputs are perceived as more valuable the more effort was ostensibly 
put into them [@Kruger2004]. Moreover, exertion of effort is deemed morally 
admirable and is rewarded, even in situations where effort does not directly 
generate additional product, quality, or economic value, suggesting that effort 
itself is moralized [@Celniker2023]. While outsourcing and reduced effort are
not synonymous, it seems the core social psychological processes are likely to
be similar: when someone outsources a task instead of doing it themselves, they
are expending less effort to achieve their goal, and people value effort. 
Indeed, some work shows that describing someone as using AI for a relational 
task led to the perception they expended less effort and were less satisfied 
with their relationship [@Liu2024]. Other work finds that people receive more 
negative social evaluations if they use AI to complete assignments at school and
at work [@Reif2025; @Roth2025]. We do not know, however, how outsourcing -- and 
therefore, presumably, expending less effort -- might be different for a human 
or an AI, and how these processes will occur in the myriad ways in which people 
can and do use AI to help them complete tasks.

Second, we need to know whether the *type* of task that people are outsourcing 
matters. One might expect outsourcing to be perceived negatively regardless of 
the type of task being outsourced -- if effort is generally moralized, then the 
domain in which it is expended (or not) should have little impact. However, 
there are reasons to expect differences between social tasks like writing vows 
and non-social tasks like writing computer code. We know that different norms,
standards, and expectations can be applied to social and non-social tasks and 
exchanges [e.g. @Fiske1992; @Heider1958; @Malle2022]. Moreover, from a 
philosophical perspective, it often matters not only *whether* something is 
done, but *how* it is done [@Aristotle2009; @Hursthouse2023; @Stohr2006]. An 
apology is not just about hearing someone say "I am sorry", but seeing genuine 
regret; a love letter is not just about hearing someone say "I love you", but 
seeing depth of emotion; and a bereavement letter is not just about hearing 
someone say "I am sorry for your loss", but seeing an understanding for the 
powerful human experience of loss. There is, perhaps especially for social 
tasks, value not only in the outcome of doing something, but the *process* too 
[@Goodman2010]. Previous work on the role of effort, however, has tended to 
focus on a limited number of tasks, looking at how people judge the quality of 
products like paintings, poems, or suits of armour based on how much effort is 
perceived [@Kruger2004] or how people judge others who expend more or less 
effort in employment and physical effort like marathon running [@Celniker2023]. 
Yet the number of tasks that people can outsource –- and therefore, presumably, 
expend less effort –- is vast, and these tasks can vary in different ways. 
Different kinds of tasks that people can outsource can be more or less socially 
relevant, more or less consequential, and more or less likely to impact others: 
it is likely that whether someone puts effort into a socially-relevant task like 
planning a friend's birthday party or writing an apology letter would have 
stronger effects on perceived morality than whether they put effort into writing 
computer code or planning a daily schedule – yet this reminds untested. To 
understand any potential negative effects of outsourcing -- to AI or to humans 
-- we must therefore look at a broad range of non-social and social tasks, 
rather than draw broad conclusions based on a few use cases.

Third, we need to know how outsourcing may shape different kinds of social 
perceptions. People can judge others on separate dimensions of warmth and 
competence [e.g., @Abele2021; @Fiske2007] as well as on dimensions of morality 
and trustworthiness [@Goodwin2014]. It remains unclear how outsourcing to AI 
might lead to differential character judgments across these different 
dimensions. For example, @Kruger2004 focus on judgments of the output, but not 
the person, and while @Celniker2023 include the core social dimensions of 
warmth, morality, and trustworthiness, we do not know how these character 
ratings would be affected by outsourcing to an AI (as well as a human) and how 
this might differ from laziness as a distinct trait. And, most importantly, we 
do not know how outsourcing different kinds of tasks might influence different 
perceptions: does outsourcing in the social domain lead to particularly negative 
effects on warmth, morality, and trustworthiness?

Fourth, we need to know how different *ways* of outsourcing to AI influence 
negative perceptions. Someone who "fully" outsources a task to AI by simply 
giving it a prompt and copying the output word-for-word might be perceived very
differently to someone who gives the AI a prompt, revises the work accordingly,
and finishes it themselves -- using AI as a *collaborative tool*, rather than as
a replacement. Similarly, someone could deceive others about their use of AI or
be perfectly honest about it, and someone could use AI because they are in a 
rush or because the task is important to them and they want to get it right. 
While it seems reasonable to assume that "fully" outsourcing would be perceived 
worse than using AI as a collaborative tool, and that not acknowledging AI use 
would be perceived worse than being honest about it, it remains unclear how much 
this reduces negative perceptions: if someone uses AI in the "best" way (by 
using it as a collaborative tool, being open about their use of AI, and having a 
good reason for using AI), would they still suffer negative social consequences 
from doing so?

Fifth, we need to understand *why* outsourcing to AI, and therefore expending 
less effort, might have these effects. Previous work has focused on how 
expending less effort directly leads to negative perceptions of others 
[@Celniker2023]. But this raises the question of *why* effort is seen as 
important and what exactly it is signalling to others, beyond one's general 
cooperative intent. It is possible that outsourcing leads to negative 
perceptions because the lack of effort spent on the task signals something more 
fundamental about how authentic one is and how much one cares about the task: 
when someone chooses to outsource a love letter to an AI, they might be seen as 
valuing that love letter and what it represents less. It could be this 
second-step order of perceptions that is the key driver of negative perceptions, 
and this, in turn, could especially be so for the kinds of tasks that are more 
socially relevant.

@@

In this paper, we build on classic social psychological work on character 
inferences from reduced effort to understand how people view others who
outsource different kinds of tasks, in different ways, for different reasons, to
AI. Across six pre-registered experiments with British participants, we seek
not only to understand how reduced effort through AI-outsourcing might shape
perceptions of others, but also advance previous theoretical work on the
moralization of effort [@Celniker2023]. Moving beyond a focus on how effort 
itself shapes character judgments, we seek to provide deeper theoretical nuance 
into why effort is important: not only because effort itself is intrinsically 
moralized, but that expending effort gives rise to second-order inferences that 
independently influence character judgments.

In Study 1, we explore the perceptions of people who outsource a range of tasks 
to AI or another person, compared to when they complete the tasks by themselves.
In Study 2, we then look at the effects of task type, AI use, and honesty. We 
explore how people perceive others who outsource different kinds of tasks with 
different levels of social relevance (e.g., from daily schedules, computer code 
and dinner recipes to wedding vows, apology letters, or bereavement cards), 
manipulating whether people use AI as a collaborative tool or "fully" outsource 
to AI and whether they are honest or deceptive about their use of AI. After 
turning to look at perceptions of both outsourcers and the outsourced work in 
Study 3, in Studies 4-6 we probe why outsourcing may have negative effects on 
how we evaluate others. In Study 4, we test potential mechanisms of perceived 
effort and authenticity by looking at how people evaluate others who either 
spend a lot or little time crafting the AI prompts, and who either outsource to 
a generic or personalized AI. In Study 5, we test the potential mechanism of 
perceived importance in the task by manipulating people's reasons for using AI 
-- either because they wanted to save time or because they cared about the task 
and thought that AI would improve their work. Finally, in Study 6, we bring 
these different potential mechanisms together to explore the different pathways 
that influence the relationship between outsourcing and negative perceptions,  
focusing on perceived effort, authenticity, and care in the task.

@@

We conducted a power simulation to determine our target sample size. The
simulation suggested that a sample size of 150 participants per condition
(overall *n* = 450 for three conditions) would be required to detect a small
difference between conditions (Cohen's *d* ≈ 0.20) with above 80% power.

```{r}
study1_sample <-
  tar_read(study1_data) %>%
  group_by(id) %>%
  summarise(
    gender = unique(gender),
    age = unique(age),
    chatgpt_used = unique(chatgpt_used)
    )
```

We recruited a convenience sample of 500 participants from the United Kingdom
through the online platform Prolific (<https://www.prolific.com/>). After 
excluding participants who failed our pre-treatment attention check, we were 
left with a final sample of `r nrow(study1_sample)` participants 
(`r sum(study1_sample$gender == "Female")` female; 
`r sum(study1_sample$gender == "Male")` male;
`r sum(study1_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study1_sample$gender, "Prefer") | is.na(study1_sample$gender))`
undisclosed gender; mean age = `r round(mean(study1_sample$age), 2)` years).
`r round(mean(study1_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before.

We randomly allocated participants into one of three conditions in a
between-subjects design: (*i*) the control condition, (*ii*) the AI outsourcing
condition, or (*iii*) the human outsourcing condition. These conditions
determined how scenarios were presented to participants.

We presented participants with six scenarios. Each scenario described a person
completing a task, such as writing computer code or writing a love letter. The
six tasks were randomly drawn from a larger set of 20 tasks (see @supptbl-tasks
for the full list of tasks). For each scenario, we told participants:

- *Control condition*: "In order to complete this task, [the person] works on it
by themselves from start to finish."
- *AI outsourcing condition*: "In order to complete this task, [the person] gets
the AI tool ChatGPT to do it for them."
- *Human outsourcing condition*: "In order to complete this task, [the person]
gets someone else to do it for them."

<br>

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package. We modelled each character evaluation --
competence, warmth, morality, laziness, and trustworthiness -- as a separate
response variable and included fixed effects for conditions, varying intercepts
for participants, and varying intercepts and slopes for tasks. We used
regularizing priors for all parameters to impose conservatism on parameter
estimates. All models converged normally ($\hat{R}$ ≤ 1.01).

For all studies in this paper, we report how we determined our sample size, all
data exclusions, all manipulations, and all measures in the studies. All studies
were pre-registered. Analyses for all studies were conducted in R v4.4.2 
[@RCoreTeam]. Visualizations were produced using the *ggplot2* and *patchwork* 
packages [@Wickham2016; @Pedersen2025]. The manuscript was reproducibly 
generated using the *targets* package [@Landau2021] and *quarto* [@Allaire2024].
All survey materials, data, and code to reproduce the analyses and figures in 
this paper can be found here: 
<https://osf.io/xhmzk/?view_only=a4da193574d7410ba4d2aa3945a28b05>

We found that people who outsourced tasks to AI or other humans were perceived
more negatively than people who completed the tasks themselves
(@fig-treatments-study1). In particular, people who outsourced were perceived as 
lazier and less competent, with smaller yet detectable differences for 
perceptions of warmth, morality, and trustworthiness
(@tbl-treatment-diffs-study1). Across all measures, outsourcing to other humans 
was perceived more negatively than outsourcing to AI.

```{r, fig.height=4, fig.width=6}
#| label: fig-treatments-study1
#| fig-cap: Overall Character Evaluations in Study 1
#| fig-align: center
#| apa-note: Participants in the control condition, the AI outsourcing condition, and the human outsourcing condition evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Colored points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_study1)
```

```{r}
#| label: tbl-treatment-diffs-study1
#| tbl-cap: Overall Pairwise Contrasts in Study 1 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on a 7-point Likert scale, pooling over participants and tasks. Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

tar_read(table_treatment_diffs_study1) %>%
  flextable() %>%
  theme_apa() %>%
  width(width = c(1.5, rep(1.5, 5))) %>%
  add_header_row(values = c("", "Response"), colwidths = c(1, 5)) %>%
  align(i = 1, j = NULL, align = "center", part = "header") %>%
  align(i = NULL, j = 1, align = "left", part = "all") %>%
  line_spacing() %>%
  fontsize(size = 10)
```

We found that the effects of outsourcing varied across the different tasks,
especially for perceptions of warmth and morality 
(@fig-treatments-tasks-study1). For example, people were perceived as
less warm if they outsourced writing a love letter, but not if they outsourced
writing computer code. Similarly, people were perceived as less moral if they
outsourced writing an apology letter to a friend, but not if they outsourced
writing a dinner recipe. By contrast, the effects of outsourcing on competence,
laziness, and trustworthiness were more consistent across tasks.

```{r, fig.height=5, fig.width=8.5}
#| label: fig-treatments-tasks-study1
#| fig-cap: Variation in the Effects of Outsourcing across Tasks in Study 1 \vspace{-30pt}
#| fig-align: center
#| apa-note: Tasks are ordered from most social (top) to least social (bottom) according to ratings from the pilot study. Point ranges are differences in marginal means on a 7-point Likert scale for the AI outsourcing condition (red) and the human outsourcing condition (blue) compared to the control condition. Points and ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_tasks_study1)
```

To determine the factors that predict variation across tasks, we incorporated
ratings of tasks from a pilot study (see Supplementary Materials). Participants 
were asked to rate the 20 tasks on several features: whether the task is social, 
requires social skills, impacts others, has important consequences, and requires 
effort. All of these features predicted stronger causal effects of outsourcing 
compared to control (Supplementary Figures [-@suppfig-interactions-study1] 
and [-@suppfig-interaction-pars-study1]). In other words, outsourcing to AI 
or other humans was perceived more negatively for tasks that were rated as being 
socially relevant and impactful.

In Study 1, we conducted our first exploration of how outsourcing to either a 
human or an AI shaped character perceptions across a variety of tasks. Our 
results show that people who outsourced the task instead of completing it 
themselves were seen as less moral, warm, competent, and trustworthy, and more 
lazy. We found that outsourcing a task led to negative character judgments 
regardless of whether the task was outsourced to a human or a machine, though 
predictably we found that outsourcing to a human was seen as worse given that 
LLMs are at least purposefully designed to be used. And finally, we found that 
the kind of task matters: these negative perceptions were not uniform, but were 
particularly found for tasks that were more social and impactful.

In Study 2, we turned to focus specifically on the different ways that people 
might expend less effort by outsourcing to an AI, focusing on the role of type 
of use (using AI as a tool but still finishing the task oneself vs. full 
delegating) and honesty about the use of AI. Would these negative perceptions be 
seen even in the "best case" scenario where someone uses AI as a tool and is 
honest about it?

@@

We recruited a convenience sample of 800 participants from the United Kingdom 
through Prolific. After excluding participants who failed our pre-treatment 
attention check, we were left with a final sample of `r nrow(study2_sample)` 
participants (`r sum(study2_sample$gender == "Female")` female; 
`r sum(study2_sample$gender == "Male")` male;
`r sum(study2_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study2_sample$gender, "Prefer") | is.na(study2_sample$gender))`
undisclosed gender; mean age = `r round(mean(study2_sample$age), 2)` years).
`r round(mean(study2_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before.

@@

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package [@Burkner2017]. We modelled each character 
evaluation -- competence, warmth, morality, laziness, and trustworthiness -- as 
a separate response variable and included fixed effects for conditions, varying 
intercepts for participants, and varying intercepts and slopes for tasks. We 
used regularizing priors for all parameters to impose conservatism on parameter 
estimates. All models converged normally ($\hat{R}$ ≤ 1.01).

@@

We first looked at the overall results averaging over tasks. Across all five 
character evaluations, we found that fully outsourcing to AI (i.e., copying the 
AI output verbatim) was perceived more negatively than using AI as a 
collaborative tool (@fig-treatments-study2; @tbl-treatment-diffs-study2). By 
contrast, we found that deception about AI usage had specific negative effects 
on perceptions of morality and trustworthiness: people who did not acknowledge 
their use of AI were perceived as less moral and less trustworthy. We did not 
find any interaction effects between full outsourcing and deception.

```{r, fig.height=6, fig.width=6}
#| label: fig-treatments-study2
#| fig-cap: Overall Character Evaluations in Study 2
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Colored points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.
